---
title: "Regression Models Course Project"
author: "Chris van Hasselt"
date: "June 21, 2015"
output: pdf_document
geometry: margin=.75in
---

add something


```{r loadLibaries,echo=FALSE,}
library(UsingR)
data("mtcars")
```

# Executive Summary

This report analyzes the relationship between miles per gallon (MPG) and an automobile's type of transmission, using the *mtcars* dataset.  Specifically, the report analyzes two questions:

#. Is an automatic or manual transmission better for MPG?
#. What is the quantified MPG difference between automatic and manual transmissions?  

# Methodology

The *mtcars* dataset consists of rows of data on different makes and models of cars, each type of car represented by a single row.  The data has two relevant columns for consideration, the *am* column and the *mpg* column.  *mpg* is the rated miles per gallon for a car; *am* takes the value 0 for automatic, or 1 for manual.

Considering the *mpg* as the outcome variable, the analysis uses the transmission type as a predictor.  With only two values, the *am* variable, while it is numeric, is essentially a factor variable.  

We would expect no difference in average mileage between cars with automatic transmissions and cars with manual transmissions.  That is, if the mean mileage for cars with manual is $\bar{x}_{M}$ and the mean mileage for cars with automatic transmission is $\bar{x}_{A}$, then 

$$
H_0: \bar{x}_{A} = \bar{x}_{M} and H_a: \bar{x}_{A} \neq \bar{x}_{M}
$$

Further, we want to investigate whether any difference in average *mpg* can be attributed to transmission type.



```{r loadLibaries,echo=FALSE,}
library(UsingR)
data("mtcars")
```


