---
title: "Regression Models Course Project"
author: "Chris van Hasselt"
date: "June 21, 2015"
output: pdf_document
geometry: margin=.75in
---


```{r loadLibaries,echo=FALSE,}
# load libraries
require(datasets)
require(stats)
require(graphics)

# specify dataset mtcars
data("mtcars")
```

# Executive Summary

This report analyzes the relationship between miles per gallon (MPG) and an automobile's type of transmission, using the *mtcars* dataset.  Specifically, the report analyzes two questions:

#. Is an automatic or manual transmission better for MPG?
#. What is the quantified MPG difference between automatic and manual transmissions?  

With some consideration of automotive mechanics, three different models are considered.  There is an MPG difference between automatic and manual transmission, but once a car's weight is taken into consideration, the difference is neglible.

# Exploratory Analysis

The *mtcars* dataset consists of rows of data on different makes and models of cars, each row representing a different type of car.  The two relevant columns to consider are the *am* column and the *mpg* column.  *am*, the transmission type, takes the value 0 for automatic, or 1 for manual; *mpg* is the rated miles per gallon, and is the outcome variable.

We want to investigate whether any difference in average *mpg* can be attributed to transmission type.  We could assert a null hypothesis of no difference in average mileage between cars with automatic transmissions and cars with manual transmissions.  That is, if the mean mileage for manual transmission is $\bar{x}_{M}$ and the mean mileage for automatic transmission is $\bar{x}_{A}$, then we could formulate a null and alternate hypotheses as

$$
H_0: \bar{x}_{A} = \bar{x}_{M}, and H_a: \bar{x}_{A} \neq \bar{x}_{M}
$$

$\bar{x}_{M} = 17.15$ and $\bar{x}_{A} = 24.39$ are different, but a simple application of a t-test is not appropriate; that type of analysis assumes the compared groups are otherwise equivalent.  The two groups have many differences, with other variables potentially explaining any difference in *mpg*.  How do you separate out the effect of just the transmission type, when we rationally know other factors also affect mileage?

# Methodology 
*qsec*, a car's quarter-mile time starting from a full stop, can be disregarded from models examining MPG, as it is an outcome variable.

Pairwise plots of *mpg* with other variables reveal the complexities in the data.  We can see a what looks like a difference between manual and automatic transmissions, with *mpg* increasing for manual transmissions.  But there is a noticeable downward trend when considering *mpg* and *wt*, *disp*, and *carb*.  It is not clear what is happening with *drat* or *gear*.  The number of gears and the rear axle ratio are key attributes of the drive train, so we would expect some combination of these to affect gas mileage.  [see Figure 1, appendix]

Considering automotive mechanics, it is reasonable to guess horsepower (*hp*), a measurement of engine performance, is related to the engine displacement (*disp*), number of carburetors (*carb*), and number of cylinders (*cyl*).  Indeed, *hp*, *carb*, *disp* and *cyl* are all correlated positively, all negatively correlated with *mpg*.  Moreover, among these power-plant measures, the strongest _negative_ correlation with *mpg* is *cyl*.  Thus, in the interest of simplifying the model, *cyl* will be used as a proxy for *hp*, *carb* and *disp*. [see appendix]

It would also be reasonable to consider the relationship between the transmission (*am*) and the number of forward gears (*gear*), as these mechanically related.  Moreover, taking account of their mechanical interaction, it makes sense to consider a model with interaction between these two variables.  

Finally, a reasonable comparison of transmission type should adjust for the weight of the vehicle, *wt*.  


```{r correlates, echo=FALSE,results='hide'}
# correlation of horsepower with displacement, cylinders
cor(mtcars$hp,mtcars$disp)
cor(mtcars$hp,mtcars$cyl)

# correlation of mpg with horsepower, cylinders, displacement
cor(mtcars$mpg,mtcars$hp)
cor(mtcars$mpg,mtcars$cyl)
cor(mtcars$mpg,mtcars$disp)

# correlation of cyl with disp
cor(mtcars$cyl,mtcars$disp)
```
# Multivariate Model

Supposing a linear model relating *mpg* and *am*, 

$$
Y_{mpg} = \beta_0 + \beta_{1}X_am + \epsilon
$$

we could look at the difference in MPG estimates for the *am* factors and see a marked difference between transmission types. 

```{r amOnlyModel,echo=FALSE}
# basic model with just am considered
mtFit0 <- lm(mpg ~ factor(am) - 1,mtcars)
summary(mtFit0)$coef
```

The difference persists when we add in the the number of cylinders as a factor variable, but is less dramatic.  
```{r amCylModel,echo=FALSE}
# basic model, with cyl added in 
mtFit1 <- lm(mpg ~ factor(am) + factor(cyl) - 1,mtcars)
summary(mtFit1)$coef
```
But with *wt* as part of our model, the difference in transmission types disappears.  
```{r amCylWtModel,echo=FALSE}
# basic model, with cyl added in 
mtFit2 <- lm(mpg ~ factor(am) + factor(cyl) + wt - 1,mtcars)
summary(mtFit2)$coef
```
# Conclusion
When gas mileage is a concern when choosing between cars, the transmission type is less important than the weight of the car.  For cars of a similar weight, the transmission type will affect gas mileage, but the difference is not as significant as the difference attributable to the weight of the car.  In short, choosing a smaller car has more of an effect on gas mileage than choice of transmission.  Once weight is taken into account, the difference is almost non-existent.

\pagebreak

# Appendix: Plots and Code

## Correlation

```{r corrCode, ref.label="correlates"}

```

## Plots of variable relationships
```{r plot,echo=FALSE, results='hide'}
pairs(mpg ~ am + wt + disp + gear+ carb + drat + cyl +hp,
      data=mtcars,
      main="MT Cars Data Exploration")
```

## Code for different models

```{r amOnlyModelCode,ref.label="amOnlyModel"}
# basic model with just am considered
mtFit0 <- lm(mpg ~ factor(am) - 1,mtcars)
summary(mtFit0)$coef
```

```{r amCylModelCode,ref.label="amCylModel"}
# basic model, with cyl added in 
mtFit1 <- lm(mpg ~ factor(am) + factor(cyl) - 1,mtcars)
summary(mtFit1)$coef
```

```{r amCylWtModelCode,ref.label="amCylWtModel"}
# basic model, with cyl added in 
mtFit2 <- lm(mpg ~ factor(am) + factor(cyl) + wt - 1,mtcars)
summary(mtFit2)$coef
```


