---
title: "Regression Models Course Project"
author: "Chris van Hasselt"
date: "June 21, 2015"
output: pdf_document
geometry: margin=.75in
---


```{r loadLibaries,echo=FALSE,}
# load libraries
require(datasets)
require(stats)
require(graphics)

# specify dataset mtcars
data("mtcars")
```

# Executive Summary

This report analyzes the relationship between miles per gallon (MPG) and an automobile's type of transmission, using the *mtcars* dataset.  Specifically, the report analyzes two questions:

#. Is an automatic or manual transmission better for MPG?
#. What is the quantified MPG difference between automatic and manual transmissions?  

# Exploratory Analysis

The *mtcars* dataset consists of rows of data on different makes and models of cars, each row representing a different type of car.  The two relevant columns to consider are the *am* column and the *mpg* column.  *am*, the transmission type, takes the value 0 for automatic, or 1 for manual; *mpg* is the rated miles per gallon.

We want to investigate whether any difference in average *mpg* can be attributed to transmission type.  We could assert a null hypothesis of no difference in average mileage between cars with automatic transmissions and cars with manual transmissions.  That is, if the mean mileage for cars with manual is $\bar{x}_{M}$ and the mean mileage for cars with automatic transmission is $\bar{x}_{A}$, then we could formulate a null and alternate hypotheses as

$$
H_0: \bar{x}_{A} = \bar{x}_{M}, and H_a: \bar{x}_{A} \neq \bar{x}_{M}
$$

$\bar{x}_{M} = 17.15$ and $\bar{x}_{A} = 24.39$ are different, but a simple application of a t-test is not appropriate; that type of analysis assumes the compared groups are otherwise equivalent.  The two groups have many differences, with other variables potentially explaining any difference in *mpg*.  How do you separate out the effect of just the transmission type, when we rationally know other factors also affect mileage?

Considering automotive mechanics, it is reasonable to assume horsepower (*hp*), a measurement of engine performance, is related to the engine displacement, number of carburetors, and number of cylinders.  Indeed, *hp*, *carb*, *disp* and *cyl* are all correlated positively; *hp*, *carb*, *cyl*, and *disp* are all negatively correlated with *mpg*.  The number of cylinders *cyl* is highly correlated with *disp*.  Moreover, among these power-plant measures, the strongest _negative_ correlation with *mpg* is *cyl*.  Thus, in the interest of simplifying the model, *cyl* will be used as a proxy for *hp*, *carb* and *disp*.  



```{r correlates, echo=FALSE}
# correlation of horsepower with displacement, cylinders
cor(mtcars$hp,mtcars$disp)
cor(mtcars$hp,mtcars$cyl)

# correlation of mpg with horsepower, cylinders, displacement
cor(mtcars$mpg,mtcars$hp)
cor(mtcars$mpg,mtcars$cyl)
cor(mtcars$mpg,mtcars$disp)

# correlation of cyl with disp
cor(mtcars$cyl,mtcars$disp)
```

Pairwise plots of *mpg* with other variables reveal the complexities in the data.  We can see a what looks like a difference between manual and automatic transmissions, with *mpg* increasing for manual transmissions.  But there is a noticeable downward trend when considering *mpg* and *wt*, *disp*, and *carb*.  It is not clear what is happening with *drat* or *gear*.  The number of gears and the rear axle ratio are key attributes of the drive train, so we would expect some combination of these to affect gas mileage.  [see Figure 1, appendix]

There are some variables from the dataset that are result measurements or 

# Multivariate Model

Supposing a linear model relating *mpg* and *am*, 

$$
Y_{mpg} = \beta_0 + \beta_{1}X_am + \epsilon
$$
\pagebreak

# Appendix: Plots and Code
```{r plot,echo=FALSE, results='hide'}
pairs(mpg ~ am + wt + disp + gear+ carb + drat + cyl +hp,
      data=mtcars,
      main="MT Cars Data Exploration")
```


With only two values, the *am* variable, while it is numeric, is essentially a factor variable. 

