---
title: "Regression Models Course Project"
author: "Chris van Hasselt"
date: "June 21, 2015"
output: pdf_document
geometry: margin=.75in
---


```{r loadLibaries,echo=FALSE,}
# load libraries
require(datasets)
require(stats)
require(graphics)

# specify dataset mtcars
data("mtcars")
```

# Executive Summary

This report analyzes the relationship between miles per gallon (MPG) and an automobile's type of transmission, using the *mtcars* dataset.  Specifically, the report analyzes two questions:

#. Is an automatic or manual transmission better for MPG?
#. What is the quantified MPG difference between automatic and manual transmissions?  

# Methodology

The *mtcars* dataset consists of rows of data on different makes and models of cars, each type of car represented by a single row.  The data has two relevant columns for consideration, the *am* column and the *mpg* column.  *mpg* is the rated miles per gallon for a car; *am* takes the value 0 for automatic, or 1 for manual.

We want to investigate whether any difference in average *mpg* can be attributed to transmission type.  We could assert a null hypothesis of no difference in average mileage between cars with automatic transmissions and cars with manual transmissions.  That is, if the mean mileage for cars with manual is $\bar{x}_{M}$ and the mean mileage for cars with automatic transmission is $\bar{x}_{A}$, then we could formulate a null hypothesis and alternate hypothesis as

$$
H_0: \bar{x}_{A} = \bar{x}_{M} and H_a: \bar{x}_{A} \neq \bar{x}_{M}
$$

But other dataset variables could explain any difference in *mpg*.  For example the cars weight (*wt*), engine displacement (*disp*), the rear axle ratio (*drat*), horsepower (*hp*), number of carburetors (*carb*), number of cylinders (*cyl*), or the number of gears (*gear*) all contribute to overall car performance.  How do you separate out the effect of just the transmission type?  

# Exploratory Analysis

Pairwise plots of *mpg* with these other variables reveal the complexities in the data.  We can see a what looks like a difference between manual and automatic transmissions, with *mpg* increasing for manual transmissions.  But there is a noticeable downward trend when considering *mpg* and *wt*, *disp*, and *carb*.  It is not clear what is happening with *drat* or *gear*.  The number of gears and the rear axle ratio are key attributes of the drive train, so we would expect some combination of these to affect gas mileage.  [see Figure 1, appendix]

Looking at the coefficients of the linear model, the estimates suggest that the type of transmission has a positive effect on *mpg* with automatic transmissions improving mileage.  But we also see that *wt* has a strong negative pull on *mpg*, with the implication that for cars of equal weight, an automatic transmission might give better gas mileage, but choosing a smaller car would save more gas.

# Multivariable Model

Supposing a linear model between taking into account all variables

\newpage
# Appendix: Plots and Code
```{r plot,echo=FALSE, results='hide'}
pairs(mpg ~ am + wt + disp + gear+ carb + drat + hp + cyl,
      data=mtcars,
      main="MT Cars Data Exploration")
```


With only two values, the *am* variable, while it is numeric, is essentially a factor variable. 

